---
title: "Analysis of Katarapko Tree Condition Data"
author:
- Department for Environment and Water
- Environmental Science and Information Branch
date: "`r format(Sys.time(), '%A, %d %B, %Y')`"
output:
  bookdown::gitbook:
    split_by: chapter
    toc_depth: 3
    css: style.css
    keep_md: yes
csl:                          "common/BibStyle.csl"
bibliography:                 ["common/refs.bib","common/packageCitations.bib"]
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, warning = FALSE, message = FALSE, echo = FALSE, results = "hide"}

# Load required packages in a manner enabling automatic citations, including base R

  packages <- c("base"
                ,"knitr"
                ,"bookdown"
                ,"readr"
                ,"readxl"
                ,"dplyr"
                ,"tidyr"
                ,"purrr"
                ,"tibble"
                ,"stringr"
                ,"forcats"
                ,"ggplot2"
                ,"lubridate"
                ,"ggridges"
                ,"gridExtra"
                ,"DT"
                ,"tmap"
                ,"DataExplorer"
                ,"rstan"
                ,"rstanarm"
                ,"mgcv" # referenced in geom_smooth
                )

  purrr::walk(packages,library,character.only=TRUE)
  
  write_bib(packages,file="common/packageCitations.bib",tweak=TRUE)
  
  
# Load useful functions
  
  source("common/functions.R")

# Set default chunk options (can adjust individual chunks differently if required)

  knitr::opts_chunk$set(warning = FALSE
                        , message = FALSE
                        , echo = FALSE
                        )
  
  options(knitr.kable.NA = "")
  
# Set mapping defaults
  
  tmap_options(basemaps = c("OpenStreetMap.Mapnik"
                            , "Esri.WorldImagery"
                            )
               )


#ggplot options
  theme_set(theme_grey()) # need to reset this to default after loading bayesplot
  
  
# don't use scientific notation for numbers
  options(scipen = 999)
  

# Other constants
  
  project <- "fpTrees"
  
```

```{r functions}

  # Tree colours
  treeCol <- c("darkred", "black", "#33A02C")
  names(treeCol) <- c("RRG","BB","Cooba")
  treeColFill <- scale_fill_manual(name = "Species",values = treeCol)
  treeColCol <- scale_colour_manual(name = "Species",values = treeCol)

```

```{r data}

  source("010_data_get.R")
  source("020_data_prep.R")
  source("030_analysis_condition.R")
  
  mapBB <- st_bbox(dat)
  
```



# Introduction



## Strategic context



## This study

Are trees/transects from Group 3 in significantly better condition than trees/transects in Group 2 in (a) 2015 and (b) 2018?

Are trees/transects from Group 3 in significantly better condition than trees/transects in Group 5 (a) 2015 and (b) 2018?
Group 1 vs Group 2/3?, Group 0 to Group 2/3?

Is condition more varied in Group 2 than Group 3, in (a) 2015 and in (b) 2018?

Is condition more varied in Group 2 than Group 5, in (a) 2015 and in (b) 2018?

    
# Methods

## Data collection

##




## Workflow

All data analysis and report writing was done in a single scripted workflow (script file: [`r list.files(here::here(),pattern="Report.Rmd",recursive=TRUE)`](`r paste0("https://github.com/Acanthiza/",project,list.files(here::here(),pattern="t.Rmd",recursive=TRUE))`) using the programs 'R' and 'R-studio'.

[R](https://www.r-project.org/) [@R-base] is an open-source platform which makes available a library of packages that can be used and modified as necessary. [R-studio](https://www.rstudio.com/) provides a range of user-friendly features to facilitate interaction with R. The packages used are listed in the appendix: [R packages used] Table \@ref(tab:packages).

All data, code and outputs are stored in a version control system at [`r project`](`r paste0("https://github.com/Acanthiza/",project,list.files(here::here(),pattern="t.Rmd",recursive=TRUE))`).

## Model

Trend in `r Y` was analysed using a Bayesian generalised linear mixed model. The analysis was run using the rstanarm package [@RN4481] in R [@R-base]. Transect was treated as a random effect in the analysis. Each tree was then assumed to provide an independent data point for the analysis. A time field was generated as $time = min(Year) + (max(Year) - min(Year))/2$. The model specification was:

``r format(formula(mod))``

## Trees

There were `r nrow(datExp)` trees available for analysis (see table below). A map of the Katarapko floodplain including tree locations is shown in Figure \@ref(fig:treesMap).

```{r trees}

  trees <- datExp %>%
    dplyr::mutate(Transect = as.factor(Transect)) %>%
    dplyr::group_by(Species,Transect,HydroGp,Year) %>%
    dplyr::summarise(n = n()
                     , UQ(rlang::sym(Y)) := median(!!ensym(Y))
                     ) %>%
    dplyr::group_by(Species,HydroGp,Year) %>%
    dplyr::summarise(Transects = n()
                     , Trees = sum(n)
                     , `Median value` = median(!!ensym(Y))
                     ) %>%
    dplyr::ungroup()

  datatable(trees
            , filter = "top"
            , extensions = c("Responsive","Buttons")
            , options = list(dom = 'Bfrtip'
                             , buttons = c('copy'
                                           #, 'csv'
                                           #, 'excel'
                                           #, 'pdf'
                                           #, 'print'
                                           )
                             )
            )

```

---

```{r treesMap, fig.cap = "Tree locations on Katarapko floodplain"}

  tmap_leaflet(
    
    tm_shape(dat) +
      tm_dots(col = "HydroGp"
              , popup.vars = c(Y,"Species","HydroGp","LOCCOMM")
              ) +
      tm_view(bbox = mapBB) +
      tm_scale_bar()
    
  )

```

---

# Results

## Diagnostics

### MCMC diagnostics

Figure \@ref(fig:mixing) and Figure \@ref(fig:rhat) provide visualisations for Markov Chain Monte Carlo (MCMC) diagnostics after fitting a Bayesian model.

```{r mixing, fig.cap = "Chain mixing. With good mixing there should be no pattern discernable"}

  stan_trace(mod)

```

---

```{r rhat, fig.cap = "Rhat values - a measure of all chains being at the same equilibrium value. These should all be close to 1."}

  plot(mod, "rhat_hist")

```

---

### Model diagnostics

Figure \@ref(fig:yResid) shows residuals plotted against `r Y`. Ideally these residuals will show no pattern - i.e. randomly distributed about zero. However, in this case, at low values of `r Y` the model is under-predicting (negative residuals) while at high values the model is over-predicting (positive residuals).

```{r yResid, fig.cap = "Residuals"}

  # Y vs. residuals
  ggplot(modResid, aes(!!ensym(Y),residual,label=Transect,colour=Transect)) +
    geom_text(size = 2
              #, alpha=0.5
              , position=position_jitter(width=0.3,height=0.2)
              ) +
    geom_hline(aes(yintercept = 0),linetype = 2) +
    facet_grid(Species~HydroGp) +
    scale_colour_viridis_c()

```

---

Figure \@ref(fig:modFit) shows how a subset of the model runs (light blue lines) fits to the original data (dark blue line).

```{r modFit, fig.height = 8, fig.cap = "Model fit"}

  # Model fit
    # Model fit
    a <- pp_check(mod) + coord_cartesian(xlim = c(0,quantile(mod$fitted.values,probs=0.05))) + labs(title = "0-quantile(0.1)")
    b <- pp_check(mod) + coord_cartesian(xlim = c(0,quantile(mod$fitted.values,probs=0.5))) + labs(title = "0-quantile(0.5)")
    c <- pp_check(mod) + coord_cartesian(xlim = c(quantile(mod$fitted.values,probs=0.5),max(mod$fitted.values))) + labs(title = "quantile(0.5)-max")
    d <- pp_check(mod) + labs(title = "0-max pred")
    
    grid.arrange(a,b,c,d)


```

## `r Y`

Table \@ref(tab:valueTab) shows median credible estimates for `r Y`.

Table \@ref(tab:valueCI) shows the 95% credible interval range for `r Y`.

Figure \@ref(fig:valueFig) shows the distribution of credible values for `r Y`.

```{r valueTab}

  kable(modRes %>%
          dplyr::select(Year,HydroGp, Species, modMedian) %>%
          tidyr::spread(Species, modMedian) %>%
          dplyr::arrange(HydroGp,Year)
        , caption = paste0("Median credible value for ",Y)
        )

```

---

```{r valueCI}

  kable(modRes %>%
          dplyr::select(Year,HydroGp, Species, text) %>%
          tidyr::spread(Species, text) %>%
          dplyr::arrange(HydroGp,Year)
        , caption = paste0("95% credible interval range for ",Y)
        )

```

---

```{r valueFig, fig.height = 8, fig.cap = paste0(Y, " change with Year for each species and group")}

  modPred %>%
    ggplot() +
      stat_density_ridges(aes(value,factor(Year),fill=Species)
                          , quantile_lines = TRUE
                          , quantiles = c(0.025, 0.975)
                          , panel_scaling = FALSE
                          , alpha = 0.5
                          ) +
      facet_grid(HydroGp~Species) +
      labs(subtitle = paste0("Distribution of credible values for ",Y)
           , caption = "Within the lines on each distribution represents 95% credible intervals"
           , y = "Year"
           , x = Y
           ) +
      treeColFill +
      theme(axis.text.y = element_text(vjust = 0))

```

## Difference to Group 2

Table \@ref(tab:diffTab) shows the likelihood that Group x has higher `r Y` than Group 2.

Figure \@ref(fig:diffFig) shows the distribution of credible values for the difference between Group x and Group 2.

```{r diffTab}

  kable(modDeltaRes %>%
              dplyr::select(-n) %>%
              dplyr::filter(direction == "Higher") %>%
              dplyr::mutate(prob = round(prob,1)) %>%
              dplyr::select(HydroGp,Year,Species,prob) %>%
              tidyr::spread(Species,prob) %>%
              dplyr::arrange(HydroGp,Year)
              
          , caption = paste0("Likelihood of higher ",Y," in Group x compared to Group 2")
        )

```

---

```{r diffFig, fig.height = 8, fig.cap = paste0("By 2018 there is little difference from Group 2. In Group 3 for all three species, and in Group 5 for RRG, ",Y," was higher than Group 2 prior to 2018 (much stronger for RRG than BB). However, by 2018 there is little difference in ",Y," within a species between groups")}

  ggplot(modDelta
           , aes(diff
                 , factor(Year)
                 , fill = Species
                 )
           ) +
      stat_density_ridges(quantile_lines = TRUE
                          , quantiles = c(0.025, 0.975)
                          , alpha=0.5
                          , panel_scaling = FALSE
                          ) +
      geom_vline(aes(xintercept = -1)
                 , linetype = 2
                 , colour = "red"
                 ) +
      geom_vline(aes(xintercept = 1)
                 , linetype = 2
                 , colour = "blue"
                 ) +
      facet_grid(HydroGp~Species) +
      labs(subtitle = paste0("Distribution of credible values for the difference from Group 2 ",Y)
           , caption = "Within the solid lines on each distribution represents 95% credible intervals\nWithin the dotted lines (-1 to 1) was defined as the same as Group 2\nBelow the red line was defined as less than Group 2\nAbove the blue line was defined as more than Group 2"
           , y = "Year"
           , x = "Group x - Group 2"
           ) +
      guides(fill = FALSE) +
      treeColFill +
      theme(axis.text.y = element_text(vjust = 0))

```

---

## Overall results

Overall model results are shown in Figure \@ref(fig:modRes).

```{r modRes, fig.cap = "Plot of model results along with raw data"}

  ggplot() +
    geom_text(data = datExp
              , aes(Year
                    ,!!ensym(Y)
                    , colour = Transect
                    , label = Transect
                    )
              , position = position_jitter()
              , size = 2
              ) +
    geom_line(data = modRes
              , aes(Year,modMedian)
              ) +
    geom_ribbon(data = modRes
                , aes(Year,ymin=modci90lo,ymax=modci90up,fill=Species)
                , alpha = 0.3
                ) +
    facet_grid(Species~HydroGp) +
    theme(axis.text.x = element_text(angle=90, hjust=0)) +
    scale_colour_viridis_c() +
    treeColFill

```

---

# Appendix

## R packages used

```{r packages, fig.cap = "R [@R-base] packages used to produce this report"}

  kable(tibble(Package = read_lines("common/packageCitations.bib")) %>%
    dplyr::filter(grepl("@",Package)
                  , !grepl("CiteR",Package)
                  ) %>%
    dplyr::mutate(Package = gsub("@Manual\\{|,|R-","",Package)) %>%
    dplyr::arrange() %>%
    dplyr::mutate(Citation = paste0("@R-",Package))
   , caption = "R [@R-base] packages used in the production of this report" 
   )

```

# References



